---
title: "Simulating and running HGAMs on spatio-temporal data"
author: "Adam Howes"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r}
knitr::opts_chunk$set(error = TRUE)

suppressPackageStartupMessages({
  library(RtGam)
  library(ggplot2)
  library(gamtools)
  library(sf)
  library(dplyr)
  library(patchwork)
})

set.seed(1)

theme_set(theme_minimal())
theme_update(
  panel.grid.minor = element_blank(),
  legend.position = "bottom"
)

options(ggplot2.discrete.colour = list(RColorBrewer::brewer.pal(4, "Dark2")))
options(ggplot2.discrete.fill = list(RColorBrewer::brewer.pal(4, "Dark2")))
```

**Background**:
We have developed a hierarchical generalized additive model (HGAM) to estimate $R_t$ from reporting triangle data across multiple spatial locations.
The model is implemented in the `gamtools` package and uses `mgcv` under the hood.
We have tested the model on real data, but not yet on simulated data where the true value of $R_t$ is known.

**Goal**:
This work aims to simulate data with known spatio-temporal structure and $R_t$ values, then to fit the HGAM model.
In doing so, we will validate the model and its implementation against a known ground truth.

**Outputs**:
Most directly, this work may be implemented using `testthat` into `gamtools` enabling unit testing of model components (e.g. `prepare_model_df()`) and integration testing of the model fitting pipeline.
Additionally, this notebook could serve as a vignette demonstrating the use of the HGAM model: data simulation, data preparation, model fitting, output post-processing, and visualization.

**Plan**:
I plan to approach this work by:

1. Understand `simulate_sir` function
1. Extend `simulate_sir` to generate reporting triangle data
1. Create fake counties spatial structure
1. Simulate correlation structures for the fake counties
1. Create time series of $R_t$ with some spatio-temporal structure
1. Apply `simulate_sir` to each county
1. Extract data and format for input to `gamtools::prep_model_df`
1. Run model with `gamtools::run_model()`
1. Visualise results
1. Perform checks ready for e.g. `testthat` setting

## 1.

```{r}
sir <- simulate_sir(
  Rt = 0.5 * sin(seq_len(100) / 10 * pi) + 1,
  S0 = 99000,
  I0 = 1000,
  R0 = 0,
  gamma = 0.5,
  delay_pmf = c(0.1, 0.2, 0.3, 0.4),
  k = 10,
  seed = 12345,
  date0 = as.Date("2023-01-01"),
  day_of_week = FALSE
)

ggplot(sir, aes(x = reference_date, y = value, col = parameter)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(~parameter, scales = "free")
```

## 2.

The `simulate_sir()` function currently doesn't generate reporting triangle data.
Here I will extend it to do so:

```{r}
sir$parameter |> unique()

reporting_delay_pmf <- c(0.6, 0.2, 0.1, 0.05, 0.05)

create_reporting_triangle <- function(sir, reporting_delay_pmf) {
  L <- length(reporting_delay_pmf)

  cases_df <- filter(sir, parameter == "observed_incident_cases") |>
    mutate(value = tidyr::replace_na(value, 0))

  cases_df |>
    rowwise() |>
    mutate(
      delay = list(0:(L - 1)),
      report_date = list(reference_date + 0:(L - 1)),
      cumulative_observed_cases = list(cumsum(rmultinom(n = 1, size = value, prob = reporting_delay_pmf))),
    ) |>
    tidyr::unnest(c(delay, report_date, cumulative_observed_cases)) |>
    select(reference_date, report_date, delay, cumulative_observed_cases)
}

reporting_triangle <- create_reporting_triangle(sir, reporting_delay_pmf)

ggplot(reporting_triangle, aes(x = reference_date, y = cumulative_observed_cases, col = as.factor(delay))) +
  geom_point() +
  scale_color_viridis_d() +
  labs(
    x = "Reference date",
    y = "Cumulative observed cases",
    col = "Delay"
  ) +
  theme_minimal()
```

## 3.

```{r}
create_sf_grid <- function(height, width) {
  sfc <- sf::st_sfc(sf::st_polygon(list(rbind(c(0, 0), c(width, 0), c(width, height), c(0, 0)))))
  grid <- sf::st_make_grid(sfc, cellsize = 1, square = TRUE)
  return(grid)
}

geometry <- create_sf_grid(height = 4, width = 4)

state_fips_code <- "99"
county_fips_code <- stringr::str_pad(seq_along(geometry), width = 3, pad = "0")
fips <- paste0(state_fips_code, county_fips_code)

sf <- sf::st_sf(geometry) |>
  mutate(fips = fips)

ggplot(sf) +
  geom_sf() +
  geom_sf_text(aes(label = fips)) +
  theme_void() +
  labs(title = "Mocked counties")
```

## 4.

```{r}
plot_matrix <- function(M, limits = NULL) {
  ggplot(
    reshape2::melt(M),
    aes(x = .data$Var1, y = .data$Var2, fill = .data$value)
  ) +
    labs(x = "", y = "", fill = "Value") +
    geom_tile() +
    scale_fill_viridis_c(limits = limits) +
    theme_void()
}

space_cov <- function(d, range, sill) {
  sill * exp(-d / range)
}

time_cov <- function(lag, rho) {
  rho^lag
}

reference_date <- seq.Date(
  from = as.Date("2025-01-01"),
  by = "day",
  length.out = 60
)

time_dist <- dist(reference_date) |> as.matrix()
time_cov <- time_cov(time_dist, rho = 0.985)

space_dist <- sf::st_distance(sf::st_centroid(sf)) |>
  as.matrix()

space_cov <- space_cov(
  space_dist,
  range = 1.5 * as.numeric(mean(space_dist)),
  sill = 1
)

# Choose to do independent by space
space_time_cov <- kronecker(diag(nrow(space_dist)), time_cov)

plot_matrix(time_cov, limits = c(0, 1)) + labs(title = "Temporal", fill = "Correlation")
```

```{r}
plot_matrix(space_cov, limits = c(0, 1)) + labs(title = "Spatial", fill = "Correlation")
```

```{r}
plot_matrix(space_time_cov, limits = c(0, 1)) + labs(title = "Spatio-temporal", fill = "Correlation")
```

## 5.

We will use a model like
$$
R_t \sim \mu_t + u_s + v_{st},
$$
where:

* $\mu_t$ is a smooth temporal trend
* $u_s$ is a spatially varying intercept
* $w_{st}$ is a spatio-temporal interaction

```{r}
n <- length(reference_date)

x <- seq_len(n)
mu <- 0.35 * (sin(x / (2 * pi)) + sin(x / (4 * pi))) + 0.65 + 0.0075 * x
mu_df <- data.frame(reference_date = reference_date, mu = mu)

ggplot(mu_df, aes(x = reference_date, y = mu)) +
  geom_line() +
  labs(title = "Temporal effect") +
  theme_minimal()
```

```{r}
u_sd <- 0.2
u <- MASS::mvrnorm(n = 1, mu = rep(0, nrow(space_cov)), Sigma = u_sd * space_cov)
u_df <- data.frame(fips = factor(fips, levels = fips), u = u) |>
  left_join(
    sf,
    by = "fips"
  )

ggplot(u_df, aes(fill = u, geometry = geometry)) +
  geom_sf() +
  geom_sf_text(aes(label = fips)) +
  scale_fill_viridis_c()
```

```{r}
v_sd <- 0.2
v <- MASS::mvrnorm(n = 1, mu = rep(0, nrow(space_time_cov)), Sigma = v_sd * space_time_cov)

rt_df <- tidyr::expand_grid(
  fips = factor(fips, levels = fips),
  reference_date = reference_date
) |>
  left_join(mu_df, by = "reference_date") |>
  left_join(u_df, by = "fips") |>
  dplyr::mutate(
    v = v,
    lp = mu + u + v,
    rt = 1 + 0.5 * tanh(lp)
  )

ggplot(rt_df, aes(x = reference_date, y = v, col = fips)) +
  geom_line() +
  labs(title = "Spatio-temporal effect") +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

```{r}
ggplot(rt_df, aes(x = reference_date, y = rt, col = fips)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(. ~ fips, scales = "free_y")
```

## 6.

This function simulates an SIR model in each county and generates the reporting triangle data:

```{r}
sir_reporting_fips <- function(fips) {
  rt <- rt_df |>
    filter(fips == .env$fips) |>
    arrange(reference_date) |>
    pull(rt)

  sir <- simulate_sir(
    Rt = rt,
    S0 = 99000,
    I0 = 1000,
    R0 = 1000,
    gamma = 0.25,
    delay_pmf = c(0.1, 0.2, 0.3, 0.4),
    k = 10,
    seed = 12345,
    date0 = as.Date("2025-01-01"),
    day_of_week = FALSE
  )

  delay <- c(0.6, 0.2, 0.1, 0.05, 0.025, 0.025)

  reporting_triangle <- create_reporting_triangle(
    sir,
    reporting_delay_pmf = delay
  ) |>
    mutate(fips_code = .env$fips, .before = "reference_date") |>
    mutate(
      state_fips_code = stringr::str_sub(fips_code, 1, 2),
      county_fips_code = stringr::str_sub(fips_code, 3, 5)
    ) |>
    rename(count_target = cumulative_observed_cases)

  return(reporting_triangle)
}

df <- purrr::map_dfr(fips, sir_reporting_fips)

ggplot(df, aes(x = reference_date, y = count_target, col = as.factor(delay))) +
  geom_point() +
  scale_color_viridis_d() +
  facet_wrap(~fips_code, scales = "free") +
  labs(
    x = "Reference date",
    y = "Cumulative observed daily cases",
    col = "Delay"
  ) +
  theme_minimal()
```

## 7.

```{r}
df <- df |>
  mutate(
    api_version = NA,
    count_total = 1,
    geo_value = NA,
    disease = "COVID-19",
    model = "gam",
    ts = as.integer(reference_date - min(reference_date))
  )

prep <- df |>
  gamtools::prep_model_df(report_date = max(df$report_date), nowcast_days = 8)

ggplot(prep, aes(x = reference_date, y = count_target, col = as.factor(delay))) +
  geom_point() +
  scale_color_viridis_d() +
  facet_wrap(~fips_code, scales = "free") +
  labs(
    x = "Reference date",
    y = "Cumulative observed cases",
    col = "Delay"
  ) +
  theme_minimal()

# Some dates are not included because of the approach
prep$reference_date |> unique()
```

## 8.

```{r}
n_distinct(prep$ts)

per_fips_delay <- prep |>
  group_by(fips_code) |>
  summarise(n_delay = n_distinct(effective_delay), .groups = "drop")

min_delay_per_fips <- min(per_fips_delay$n_delay)

n_pair_fips_ts <- n_distinct(interaction(prep$fips_code, prep$ts, drop = TRUE))

fit <- gamtools::fit_model(
  df = prep, network = "none"
)

summary(fit)
```

## 9.

```{r}
pred_prep <- prep |>
  prep_prediction_df(
    nowcast_days = 4,
  )

summary <- summarize_posterior_preds(pred)
```

$R_t$ values:

```{r}
delay <- c(0.6, 0.2, 0.1, 0.05, 0.025, 0.025)
gi <- c(0.1, 0.2, 0.3, 0.4)

expectations_rt <- df |>
  prep_rt_df(
    nowcast_days = 0,
    expected_delay = sum(delay * 0:(length(delay) - 1)),
    gi_len = length(gi)
  ) |>
  mutate(count_denominator = 1) |>
  add_posterior_expectations(fit)

rt <- expectations_rt |>
  add_posterior_rt(fit, gi_pmf = gi, col_name = "fips_code")

rt_summary <- summarize_posterior_rt(rt)
```

## 10.

```{r}
df_retrospective <- df |>
  group_by(reference_date) |>
  filter(delay == max(delay))

ggplot(summary, aes(x = reference_date)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), fill = "#deebf7") +
  geom_ribbon(aes(ymin = lower_50, ymax = upper_50), fill = "#9ecae1") +
  geom_line(aes(y = median), col = "#08519c") +
  geom_point(data = df_retrospective, aes(x = reference_date, y = count_target), shape = 1) +
  facet_wrap(~fips_code, scales = "free")

ggsave("forecast.png", h = 10, w = 12, bg = "white")
```

```{r}
rt_df_corrected <- rt_df |>
  mutate(fips_code = fips)

ggplot(rt_summary, aes(x = reference_date)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), fill = "#fee5d9") +
  geom_ribbon(aes(ymin = lower_50, ymax = upper_50), fill = "#fcae91") +
  geom_line(aes(y = median), col = "#cb181d") +
  geom_line(data = rt_df_corrected, aes(y = rt)) +
  facet_wrap(~fips_code, scales = "free")

ggsave("rt.png", h = 10, w = 12, bg = "white")
```
