---
title: "Introduction to RtGam"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Introduction to RtGam}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

RtGam is an R package for real-time estimation of epidemic growth rates.
The package fits a smooth trend over time to an epidemic timeseries with a generalized additive model (GAM).
From this fitted trend, it generates a short-term forecast and estimates the time-varying effective reproductive number, $R_t$ and intrinsic growth rate, $r$.

This vignette gives a brief overview of RtGam.
It walks through how to fit a model to an epidemic timeseries and interpret the results.

## Setup

This vignette assumes you have the RtGam package installed and loaded.
This vignette uses `ggplot2`.

```{r setup}
# Install with remotes::install_github("CDCGov/cfa-gam-rt")
library(RtGam)
library(ggplot2)
```

## Data

The `RtGam` package requires a time series of observed incident case counts.
For demonstration purposes, the package includes a simulated epidemic dataset, `stochastic_sir_rt`, adapted from @gostic2020practical.

In this example dataset, `obs_incidence` is the vector of observed case counts, and `date` is the corresponding vector of dates.
The dataset has a timeseries of 299 days that we'll subset to 60 days to make plotting easier.


```{r plot data}
data <- stochastic_sir_rt[41:100, ]
ggplot(data) +
  geom_point(aes(date, obs_incidence)) +
  theme_bw() +
  labs(x = "Reference date", y = "Observed cases")
```

Because this dataset was simulated, we know the true $R_t$ that generated the epidemic timeseries.
```{r plot true rt}
ggplot(data) +
  geom_point(aes(date, true_rt)) +
  theme_bw() +
  labs(x = "Reference date", y = "True Rt") +
  scale_y_continuous(
    trans = "log",
    labels = scales::number_format(accuracy = 0.01)
  )
```

For more on this dataset see `?stochastic_sir_rt`.

## Fitting the Model

To fit an `RtGam` model, provide a vector of observed case counts and their corresponding dates.
The model can automatically select hyperparameters like the smoothing basis dimension and the penalty basis dimension.

```{r fit model}
fit <- RtGam(
  cases = data[["obs_incidence"]],
  reference_date = data[["date"]]
)
print(fit)
```

To select your own smoothing or penalty basis dimension you can provide those parameters to the model.
The penalty basis dimension can be specified with the `k` argument and the smooth basis dimension can be specified with the `m` argument.
For example, we can decrease the penalty basis dimension of the model by setting `m` to 1.
See `smooth_dim_heuristic()` and `penalty_dim_heuristic()` for more on how these basis dimensions are set and when to change them.

```{r fit model again}
fit_non_adaptive <- RtGam(
  cases = data[["obs_incidence"]],
  reference_date = data[["date"]],
  m = 1
)
print(fit_non_adaptive)
```

Note that RtGam throws warnings if diagnostic issues are detected.
See `?RtGam` for more details on how to specify a model and `?check_diagnostics` for more on model diagnostics.

## Analyzing Results

### Posterior predicted cases

The `predict()` method draws samples from the posterior distribution of the fitted model.
By default it draws 100 sampled timeseries of posterior predicted cases.

```{r predict obs cases}
head(predict(fit))
```

Calling `plot` on the fitted model object also draws posterior samples and plots them against the provided timeseries of cases.

```{r plot obs cases}
plot(fit)
```

### Expected $R_t$

Producing an estimate of $R_t$ requires some additional information.
It needs the `mean_delay`, which is the expected number of days from a case's _infection_ to the case being counted.
For example in a timeseries of hospitalizations, it's the average number of days between infection and being hospitalized.

It also requires an estimate of the generation interval, which can be drawn from the literature or estimated separately using data on paired cases.
`RtGam` expects the GI to be formatted as a probability mass function.
We recommend using the package [primarycensored](https://primarycensored.epinowcast.org) to format a known generation interval distribution as a PMF.
The probability mass function should have the first element dropped and replaced with a 0 because the underlying renewal-equation based approach assumes no same-day transmission.

In this simulated example, cases are observed immediately so the mean delay is 0.
We have pre-generated the generation interval probability mass function (see more with `?sir_gt_pmf`).

```{r plot Rt}
plot(
  fit,
  parameter = "Rt",
  mean_delay = 0,
  gi_pmf = sir_gt_pmf
)
```

## Model validation

Comparing the posterior draws from the fitted model posterior to the simulated ground truth suggests that the model is able to recover $R_t$ from this timeseries.
The model is able to identify and correctly time the sharp drop in $R_t$ in early March.
However, we can see that the model has high uncertainty during initialization and struggles to adjust to the abrupt increase in $R_t$ in early April.

```{r plot Rt against data}
plot(
  fit,
  parameter = "Rt",
  mean_delay = 0,
  gi_pmf = sir_gt_pmf
) +
  geom_point(aes(date, true_rt), data = data)
```

Future development effort will aim to improve model performance on challenging datasets like this example.

## Additional Resources

To provide feedback or ask questions, refer to the package's [GitHub repository](https://github.com/CDCgov/cfa-gam-rt).

*Note: This package is in early development.
Outputs may be misleading or incorrect. Use with caution and refer to the repository for updates.*

## References
